---
title: "DEGs for All Cell Types on WT Genotype"
author: "Dario Righelli"
date: "`r Sys.Date()`"
editor_options: 
    chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, include=TRUE)
```

# Loading libraries

This chunk loads all the necessary R packages and resources for the downstream single-cell and pseudobulk analyses.

-   `scater`: used for visualization and processing of single-cell data.
-   `RUVSeq`: provides functions for normalization using control genes (RUV normalization).
-   `darioscripts`: a custom helper package with user-defined functions.
-   `DT`: used to generate interactive HTML tables.
-   `ggpubr` and `ggrepel`: utilities to produce clean ggplot-based visualizations with non-overlapping text labels.
-   `utils.R`: a script with additional helper functions, sourced from the `../R/` directory.
-   The color palette used for plotting (`subcluster.color`) is loaded from a pre-saved `.RData` file and assigned to `palette`.

Some libraries and scripts are commented out, likely because they are either not currently needed or replaced by custom implementations.

```{r}
library(scater)
library(RUVSeq)
library(darioscripts)
library(DT)
library(ggpubr)
library(ggrepel)
source("R/utils.R")
load("~/Downloads/multiome_data/data/MultiomeAllenColorLabel.RData")
palette <- subcluster.color
```

# Visualization for WT Samples

This chunk performs dimensionality reduction and visualization for wild-type (WT) and sleep-deprived (SD) samples from a filtered single-cell RNA-seq dataset.

-   The `SingleCellExperiment` object list `scelist` is first read from a file containing RNA-only data with doublets removed.
-   Only WT and SD samples are selected using a regex filter.
-   Dimensionality reduction is computed using PCA, followed by `tSNE` and `UMAP`, which are embedded into the `scelistwt` objects.
-   Sample names containing `_WT` are renamed to `_HC` (likely to represent healthy controls).
-   For each sample, two plots are created: a tSNE and a UMAP, colored by annotated cell labels.
-   These plots are arranged side-by-side and labeled with the sample name using `ggpubr`.

This step is useful for visually inspecting clustering patterns and checking label separability across conditions.

```{r}
scelist <- readRDS("~/Downloads/multiome_data/data/scelist_SR_dbl_signacfilt_dimred_noATAC_doublrm.RDS")
scelistwt <- scelist[grep("_WT|_SD$", names(scelist))]
library(scater)
set.seed(3)
scelistwt <- lapply(scelistwt, runTSNE, dimred="PCA")
scelistwt <- lapply(scelistwt, runUMAP, dimred="PCA")
names(scelistwt) <- gsub("_WT", "_HC", names(scelistwt))
lapply(seq_along(scelistwt), function(i)
{
    sce <- scelistwt[[i]]
    ts <- plotReducedDim(sce, dimred="TSNE", colour_by="labels", point_size=1) +
        scale_color_manual(values=palette) + theme_bw()
    um <- plotReducedDim(sce, dimred="UMAP", colour_by="labels", point_size=1) +
        scale_color_manual(values=palette) + theme_bw()
    gga <- ggarrange(plotlist=list(ts, um), common.legend=TRUE)
    annotate_figure(gga, top=ggpubr::text_grob(names(scelistwt)[i], 
    color = "black", face = "bold", size = 14))
})

```

# Load and Annotate Pseudobulk Data (All Cell Types)

In this section, we load pseudobulk gene expression data and perform key preprocessing steps:

-   The `multiome` object, a `SingleCellExperiment` storing pseudobulk counts, is loaded from disk.
-   Each sample is annotated with a `superclass` label: either "Non Neuronal", "Neuronal - Glutamatergic", or "Neuronal - Gabaergic" based on known cell types.
-   We load a list of known **negative control genes** (used later in RUV normalization) and assign a subset as a test set (10% of them).
-   Similarly, **positive control genes** are loaded and flagged.
-   Sample condition "CTRL" is renamed to "HC" for consistency.
-   We then restrict the data to only **wild-type (WT)** samples and remove cell types not detected in **all samples** (fewer than 6 replicates).
-   Library size normalization and log-normalization are performed.
-   Finally, we compute and plot MDS and PCA embeddings to assess clustering across cell types and conditions.

```{r}
multiome <- readRDS("~/Downloads/multiome_data/data/GEX_pseudo_2023.RDS") # pseudobulk

multiome$superclass <- "Non Neuronal"

multiome$superclass[multiome$label %in% c("CA1", "CA1-do","CA1-ProS", "CA2", "CA3", "SUB-ProS", "DG")] <- "Neuronal - Glutamatergic"
multiome$superclass[multiome$label %in% c("Lamp5", "Sncg", "Vip", "Serpinf1",
                                          "Ntng1", "Pvalb", "Sst", "Sst-Chodl")] <- "Neuronal - Gabaergic"


nc <- readRDS("~/Downloads/multiome_data/ctrls/SD_neg_ctrl.RDS")
dim(nc)
rowData(multiome)$negctl <- FALSE
rowData(multiome)$negctl[rownames(multiome) %in% nc$symbol] <- TRUE

set.seed(3)
idxsamp <- sample(seq_along(nc$ensembl_id), round((dim(nc)[1]*10)/100))
nctest <- nc[idxsamp,]
nc <- nc[-idxsamp,]
rowData(multiome)$negctl <- FALSE
rowData(multiome)$negctl[rownames(multiome) %in% nc$symbol] <- TRUE
rowData(multiome)$negctltest <- FALSE
rowData(multiome)$negctltest[rownames(multiome) %in% nctest$symbol] <- TRUE

newps <- read.table("~/Downloads/multiome_data/ctrls/111022_Positive_Controls_from_BMC_Genomics_AddFile2.txt", header=TRUE)
newpshits <- newps
rowData(multiome)$posctl <- FALSE
rowData(multiome)$posctl[rownames(multiome) %in% newps$Gene_Name] <- TRUE

newpshits$kept <- FALSE
newpshits$kept[which(newpshits$Gene_Name %in% rownames(multiome))] <- TRUE

multiome$condition[multiome$condition=="CTRL"] <- "HC"
multiomewt <- multiome[,multiome$genotype=="WT"]
noct <- names(which(table(multiomewt$label)<6))
multiomewt <- multiomewt[,-which(multiomewt$labels %in% noct)]
multiomewt <- computeLibraryFactors(multiomewt)
multiomewt <- logNormCounts(multiomewt)
multiomewt <- runMDS(multiomewt)
mdal <- scater::plotMDS(multiomewt, color_by="labels", shape_by="condition", point_alpha=1, point_size=2) +theme_bw()
pcal <- plotPCA(multiomewt, color_by="labels", shape_by="condition", point_alpha=1, point_size=2)+theme_bw()
ggarrange(plotlist=list(mdal, pcal), common.legend=TRUE)

```

# Annotate Genes with External Feature Information

In this step, gene-level metadata is added to the `rowData` of the `multiomewt` object. These annotations include:

-   Gene symbol (`TENxSymbol`)
-   Ensembl gene ID (`TENxEnsembl`)
-   Chromosome (`TENxChr`)
-   Genomic start and end positions (`TENxStart`, `TENxEnd`)

These annotations are matched using a reference file containing information on duplicated gene features (`features_duplicates.tsv`). This mapping enables downstream integration with genomic coordinates and facilitates better interpretation of differentially expressed genes.

```{r}
feat <- read.table("~/Downloads/multiome_data/reference/features_duplicates.tsv", sep="\t", header=TRUE)
rn <- rownames(rowData(multiomewt)) 
idx <- which(rn %in% feat$symbol)
idx1 <- which(feat$symbol %in% rn)
rowData(multiomewt)$TENxSymbol[idx] <- feat$symbol[idx1]
rowData(multiomewt)$TENxEnsembl[idx] <- feat$ensembl[idx1]
rowData(multiomewt)$TENxChr[idx] <- feat$chr[idx1]
rowData(multiomewt)$TENxStart[idx] <- feat$start[idx1]
rowData(multiomewt)$TENxEnd[idx] <- feat$end[idx1]
```

# Dimensionality Reduction for Each Cell Type

In this section, we explore the variation between experimental conditions (`HC` and `SD`) at the **single cell type level**.

-   We loop through each cell type (label) in the `multiomewt` dataset.
-   For each subset (corresponding to one label), we perform dimensionality reduction using both **MDS** and **PCA**.
-   The resulting plots are colored by experimental condition and shaped by replicate.
-   We generate side-by-side plots of the two reductions (MDS and PCA) for each label, with a common legend and a title showing the cell type name.

This helps visualize whether the sleep deprivation (`SD`) and control (`HC`) conditions lead to visible shifts in expression space within each cell type.

```{r}
lbls <- unique(multiomewt$labels)
bnormpl <- lapply(lbls, function(l)
{
    sce <- multiomewt[,multiomewt$labels %in% l]
    md <- scater::plotMDS(sce, color_by="condition", shape_by="replicate", point_alpha=1)+theme_bw()+scale_color_manual(values=c(SD="black", HC="red"))
    pc <- plotPCA(sce, color_by="condition", shape_by="replicate", point_alpha=1 )+theme_bw()+scale_color_manual(values=c(SD="black", HC="red"))
    gg <- ggarrange(plotlist=list(md, pc), common.legend=TRUE)
    gg <- annotate_figure(gg, top=ggpubr::text_grob(l,
                    color = "black", face = "bold", size = 14))
    return(gg)
})

names(bnormpl) <- lbls
```

# Normalization Using Upper Quartile + RUVs (per Cell Type)

In this section, we apply normalization to each cell type using a combination of **Upper Quartile (UQ)** normalization and **RUVs** (Removal of Unwanted Variation using replicate samples). This step helps reduce technical noise and unwanted batch effects.

-   The dataset `multiomewt` is split by individual cell type labels.
-   For each subset:
    -   **Upper Quartile normalization** is applied to reduce library size effects.
    -   The condition (`HC` or `SD`) is grouped to define biological replicates.
    -   The `RUVs` function is used to estimate 3 unwanted variation factors (k = 3), using negative control genes.
    -   The resulting normalized counts are saved under either `"uqruvsk3"` (if UQ was successful) or `"ruvsk3"` (if not).
    -   PCA and MDS are recomputed using the normalized data.
    -   The normalization method is saved in the metadata of the `SingleCellExperiment`.

This step ensures that downstream differential expression analysis is robust to technical variability.

```{r}
lbls <- unique(multiomewt$labels)

scelist <- lapply(lbls, function(l)
{
    multiomewt[,multiomewt$labels %in% l]
})

names(scelist) <- lbls

scelist <- lapply(scelist, function(sce)
{
    reducedDims(sce)$MDS <- NULL
    uq <- EDASeq::betweenLaneNormalization(counts(sce), which="upper")
    grp <- makeGroups(sce$condition)
    rownames(grp) <- unique(sce$condition)
    k=3
    if ( isEmpty(which(is.na(uq), arr.ind=TRUE)) )
    {
        mat <- round(as.matrix(uq))
        norm <- "uqruvsk3"
    } else {
        mat <- round(as.matrix(counts(sce)))
        norm <- "ruvsk3"
    }
    ruvs <- RUVs(x=mat, cIdx=rownames(sce)[rowData(sce)$negctl], k=k,
                     scIdx=grp, round=TRUE)
    colData(sce) <- cbind.DataFrame(colData(sce), ruvs$W)
    assays(sce)[[norm]] <- ruvs$normalizedCounts
    sce <- runMDS(sce, assay.type=norm)
    sce <- runPCA(sce, assay.type=norm)
    
    metadata(sce)$norm <- norm
    return(sce)
})
names(scelist) <- lbls

```

# Plot MDS and PCA after Normalization (per Cell Type)

This section visualizes the effect of normalization and the separation of samples by condition (`HC` vs `SD`) for each cell type individually.

-   For each cell type in the list `scelist`:
    -   We extract the normalized `SingleCellExperiment` object.
    -   We generate two dimensionality reduction plots: one using **MDS** and one using **PCA**.
    -   Both plots are colored by the experimental condition and shaped by replicate, making batch effects and biological trends visually detectable.
    -   The plots are arranged side-by-side using `ggarrange`, and a title is added that includes the cell type and the normalization method used (from `metadata(sce)$norm`).
-   Finally, for each cell type, the same set of plots is printed twice vertically using `ggarrange` with `nrow=2` (may be for layout testing or optional repetition).

```{r}
normpl <- lapply(seq_along(scelist), function(i)
{
    sce <- scelist[[i]]
    l <- names(scelist)[i]
    md <- scater::plotMDS(sce, color_by="condition", shape_by="replicate", point_alpha=1 )+theme_bw()+scale_color_manual(values=c(SD="black", HC="red"))
    pc <- plotPCA(sce, color_by="condition", shape_by="replicate", point_alpha=1 )+theme_bw()+scale_color_manual(values=c(SD="black", HC="red"))
    gg <- ggarrange(plotlist=list(md, pc), common.legend=TRUE)
    gg <- annotate_figure(gg, top=ggpubr::text_grob(
                            paste0(l, " ", metadata(sce)$norm),
                            color="black", face="bold", size=14))
    return(gg)
})

names(normpl) <- names(scelist)
for (l in lbls)
{
    print(ggarrange(plotlist=list(normpl[[l]], normpl[[l]]), nrow=2))
}
```

# Run Differential Expression with UQ+RUVsK3 Normalization

In this step, we perform **differential gene expression (DGE)** analysis for each cell type using the **edgeR** framework, incorporating normalization from the previous RUVs (Removing Unwanted Variation) step.

-   Iterates over each normalized `SingleCellExperiment` in `scelist`.
-   Constructs a `DGEList` object from raw counts and associated metadata.
-   If the sample was normalized with **Upper Quartile + RUVs (uqruvsk3)**, it computes **offsets** using library size and normalization factors to account for sequencing depth.
-   Filters lowly expressed genes using `filterByExpr()`, requiring at least 5 counts across conditions.
-   Calls a custom `applyEdgeR()` function, applying the contrast `"SD - HC"` and including the RUV W factors (`W_1`, `W_2`, `W_3`) as covariates.
-   Reorders and integrates the differential results into the `rowData()` of the original object.
-   Flags significantly differentially expressed genes (`FDR < 0.05`) with a `sign` column.
-   (Commented out) An optional block to save the resulting DEG tables for each cell type as `.tsv`

```{r}
scelist <- lapply(scelist, function(sce)
{
    y <- DGEList(counts(sce), samples=colData(sce))
    if (metadata(sce)$norm=="uqruvsk3")
    {
        y1 <- calcNormFactors(y, method="upperquartile")
        offs <- log(y1$samples$lib.size*y1$samples$norm.factors)
    } else {
        offs <- NULL
    }
    keep <- filterByExpr(y, group=sce$condition, min.count=5, min.total.count=0)
    keep.genes <- data.frame(keep[keep=="TRUE"])
    sce <- sce[rownames(keep.genes),]
    cntrs <- c("SD - HC")
    res <- applyEdgeR(counts=counts(sce), colData=colData(sce),
                          factors="condition", wnames=c("W_1", "W_2", "W_3"),
                          contrasts=cntrs, p.threshold=1,
                          edgeroffs=offs)[[1]]
    res <- res[order(match(rownames(res),rownames(sce))),]
    rowData(sce) <- rowData(sce)[,-2]
    rowData(sce) <- cbind.DataFrame(rowData(sce), res)
    rowData(sce)$sign <- rowData(sce)$FDR < 0.05
    return(sce)
})

```

# Compare DEGs and Nuclei Count

This section explores the relationship between the number of nuclei (cells) detected per cell type and the number of differentially expressed genes (DEGs) identified. This is done in two parts: using all cell types, and then filtering to those with more than 100 nuclei to reduce noise from low-abundance populations.

## Compare DEGs and Nuclei Count (All Cell Types)

For each cell type: - The function `statDEG()` summarizes DEG statistics, likely counting the number of significant genes. - These summaries are collected into a combined table with `buildTab()`, which returns a `data.frame` containing at least:\
- the number of nuclei (`Nuclei`),\
- the number of DEGs (`DEGs`),\
- and the label (`Label`) for each cell type. - A scatter plot is created: - **X-axis:** log-transformed number of nuclei. - **Y-axis:** log-transformed number of DEGs. - Labels are annotated using `geom_text_repel()` for readability. - A quadratic regression line is fitted (`y ~ poly(x, 2)`) to highlight any non-linear trends. - The results are visualized both as a `ggplot` and as an interactive data table using `DT::datatable()`.

```{r}

scelist <- lapply(scelist, function(sce)
{
    statDEG(sce)
})

tab <- buildTab(scelist)
p1 <- ggplot(tab, aes(x = log(Nuclei), y = log(DEGs), color = Label)) +
    geom_point(size = 3) +
    #scale_color_manual(values = neuronal_color) + 
    theme_classic() +
    theme(legend.position = "none", axis.text = element_text(size = 7), 
          axis.title = element_text(size = 7)) +
    stat_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, col = "grey") +
    geom_text_repel(aes(label = Label), size = 3,
                    box.padding   = 0.35) +
    ggtitle("WT")
datatable(tab)
p1
```

## Compare DEGs and Nuclei (Filtered: \>100 Nuclei)

This analysis filters out low-abundance cell types, retaining only those with **more than 100 detected nuclei**. This threshold helps focus on well-sampled populations, where differential expression analysis is statistically more reliable.

-   A subset `tab1` is created from the original summary table `tab`, keeping only cell types with more than 100 nuclei.
-   A scatter plot is generated to show the relationship between the **log-transformed number of nuclei** and **log-transformed number of DEGs**.
-   Labels for each point represent the cell type (`Label`), with overlap prevented using `geom_text_repel`.
-   A polynomial trend line (`y ~ poly(x, 2)`) is added to model the relationship.
-   The corresponding data is displayed in an interactive table using `DT::datatable`

```{r}
tab1 <- tab[tab$Nuclei>100,]
p1 <- ggplot(tab1, aes(x = log(Nuclei), y = log(DEGs),color = Label)) +
  geom_point(size = 3) +
  #scale_color_manual(values = neuronal_color) + 
  theme_classic() +
  theme(legend.position = "none", axis.text = element_text(size = 7), 
        axis.title = element_text(size = 7)) +
  stat_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, col = "grey") +
    geom_text_repel(aes(label = Label), size = 3,
                   box.padding   = 0.35)+#,
    ggtitle("WT 100 nuclei")
datatable(tab1)
p1
```

## Compare DEGs vs Nuclei in Neuronal Types (\>100 Nuclei)

This analysis focuses specifically on **neuronal populations** that are well-sampled (i.e., having more than 100 nuclei). By excluding non-neuronal cells, the relationship between sample size and DEG count is analyzed within a more functionally homogeneous group.

Steps performed: - Filter `tab` to retain only rows where the number of nuclei exceeds 100. - Further subset to include only neuronal cell types, based on the `superclass` annotation in the original `multiome` dataset. - Create a scatter plot showing the log-transformed number of nuclei vs. log-transformed number of DEGs. - Annotate cell types with `geom_text_repel` for clarity. - Add a quadratic regression line to highlight overall trends. - Present results both graphically and as an interactive table.

```{r}
tab2 <- tab[tab$Nuclei>100,]
neuronallabels <- unique(multiome$label[multiome$superclass != "Non Neuronal"])
tab2<-tab2[tab2$Label %in% neuronallabels,]
p1 <- ggplot(tab2, aes(x = log(Nuclei), y = log(DEGs), color=Label)) +
  geom_point(size = 4) +
  theme_classic() +
  theme(legend.position = "none", axis.text = element_text(size = 7), 
        axis.title = element_text(size = 7)) +
  stat_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, col = "grey") +
  geom_text_repel(aes(label = Label), size = 3,
                   box.padding   = 0.35)+
    ggtitle("WT 100 nuclei neuronal")
datatable(tab2)
p1
```

## MDS Plots of Superclasses

This section explores differences in gene expression between broad neuronal classes — **GABAergic**, **Glutamatergic**, and **Non-Neuronal** — in wild-type samples. It involves the following steps:

-   Load the pseudobulk `multiome` dataset.
-   Define three main cell-type superclasses.
-   Load and annotate both negative and positive control genes used for normalization.
-   Filter out low-represented cell types (fewer than 6 replicates).
-   Normalize each subclass and compute MDS (Multidimensional Scaling).
-   Generate separate MDS plots for each superclass subset.

```{r}
multiome <- readRDS("~/Downloads/multiome_data/data/GEX_pseudo_2023.RDS") # pseudobulk

multiome$superclass <- "Non Neuronal"

multiome$superclass[multiome$label %in% c("CA1", "CA1-do","CA1-ProS", "CA2", "CA3", "SUB-ProS", "DG")] <- "Glutamatergic"
multiome$superclass[multiome$label %in% c("Lamp5", "Sncg", "Vip", "Serpinf1",
                                          "Ntng1", "Pvalb", "Sst", "Sst-Chodl")] <-
    "Gabaergic"
nc <- readRDS("~/Downloads/multiome_data/ctrls/SD_neg_ctrl.RDS")
dim(nc)
rowData(multiome)$negctl <- FALSE
rowData(multiome)$negctl[rownames(multiome) %in% nc$symbol] <- TRUE
set.seed(3)
idxsamp <- sample(seq_along(nc$ensembl_id), round((dim(nc)[1]*10)/100))
nctest <- nc[idxsamp,]
nc <- nc[-idxsamp,]
rowData(multiome)$negctl <- FALSE
rowData(multiome)$negctl[rownames(multiome) %in% nc$symbol] <- TRUE
rowData(multiome)$negctltest <- FALSE
rowData(multiome)$negctltest[rownames(multiome) %in% nctest$symbol] <- TRUE
newps <- read.table("~/Downloads/multiome_data/ctrls/111022_Positive_Controls_from_BMC_Genomics_AddFile2.txt", header=TRUE)
newpshits <- newps
rowData(multiome)$posctl <- FALSE
rowData(multiome)$posctl[rownames(multiome) %in% newps$Gene_Name] <- TRUE
newpshits$kept <- FALSE
newpshits$kept[which(newpshits$Gene_Name %in% rownames(multiome))] <- TRUE
multiome$condition[multiome$condition=="CTRL"] <- "HC"
multiomewt <- multiome[,multiome$genotype=="WT"]
noct <- names(which(table(multiomewt$label)<6))
multiomewt <- multiomewt[,-which(multiomewt$labels %in% noct)]
multiomewt <- computeLibraryFactors(multiomewt)
multiomewt <- logNormCounts(multiomewt)
multiomewt <- runMDS(multiomewt)
gaba <- multiomewt[,multiomewt$superclass=="Gabaergic"]
gaba <- computeLibraryFactors(gaba)
gaba <- logNormCounts(gaba)
gaba <- runMDS(gaba)
scater::plotMDS(gaba, color_by="labels", shape_by="condition", point_alpha=1) + ggtitle("WT Gabaergic Neurons")+scale_color_manual(values=palette)

gluta <- multiomewt[,multiomewt$superclass=="Glutamatergic"]
gluta <- computeLibraryFactors(gluta)
gluta <- logNormCounts(gluta)
gluta <- runMDS(gluta)
scater::plotMDS(gluta, color_by="labels", shape_by="condition", point_alpha=1) + ggtitle("WT Glutamatergic Neurons")+scale_color_manual(values=palette)

nneu <- multiomewt[,multiomewt$superclass=="Non Neuronal"]
nneu <- computeLibraryFactors(nneu)
nneu <- logNormCounts(nneu)
nneu <- runMDS(nneu)
scater::plotMDS(nneu, color_by="labels", shape_by="condition", point_alpha=1) + ggtitle("WT Non Neuronal")+scale_color_manual(values=palette)
```

# SessionInfo

This final chunk prints the current R session information, which includes details about the R version, loaded packages, and the environment. This is useful for reproducibility and debugging.

```{r, tidy=TRUE}
sessionInfo()
```
