---
title: "Data Loading"
author: "Dario Righelli"
date: "`r Sys.Date()`"
editor_options: 
    chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache=FALSE)
```

# Raw data preprocessing

To process raw 10x BCL files produced by the sequencer, we take aid of
the 10x [cellranger-arc
software](https://www.10xgenomics.com/support/software/cell-ranger-arc/latest)
Firstly, we used the **makefastq** command to generate fastq files for
each sample GEX/ATAC and then with the **count** command by aid of the
**libraries** files, we combined GEX and ATAC for each sample to obtain
a unique output folder. This results in a unique output folder for each
sample.

Details of the implemented MAKEFILE(s) are present in the [main github
repository](https://github.com/RissoLab/MultiomeSD) of this website. In
particular, the files are present in the
[inst/preprocessing](https://github.com/RissoLab/MultiomeSD/tree/qwebsite/inst/preprocessing)
folder, where they are organized by replicate. Refer to the MAKEFILE
present in each of these folders to replicate the results.

# Load required libraries

This chunk loads the necessary Bioconductor packages: -
`SingleCellExperiment`: data structure for single-cell assays -
`TENxMultiomeTools`: helpers for 10x Genomics multiome data - `scater`,
`scuttle`: normalization, QC, and visualization tools

```{r}
library(SingleCellExperiment)
library(TENxMultiomeTools)
```

# Data loading

After processing the FASTQ files with Cell Ranger ARC, we are able to
load each replicate as a separate SingleCellExperiment object.

Here we show the code to use to load the data into R by using the
`TENxMultiomeTools` package.

```{r, eval=FALSE}
# After placing on the right path where all the replicates have been processed 
# and stored 
dirs <- list.dirs(".", full.names=TRUE)
repfolders <- dirs[grep("REP[^/]*?/outs$", dirs)]
scelist <- lapply(repfolders, read10xMultiome, type="HDF5")
names(scelist) <- unlist(lapply(strsplit(repfolders, split="/"), function(x){x[3]}))
```

This produces a list of sce objects with two main experiments, one, as
default, named RNA, and an `altExp` named ATAC.

# Data Splitting

To speed up further computations, and to lighten the load on memory here
each sce is split in two different objects one with only RNA and one
with only ATAC.

```{r, eval=FALSE}
scelistnoatac <- lapply(scelist, function(sce){
    altExp(sce) <- NULL
    sce
})
saveRDS(scelistnoatac, file="~/Downloads/multiome_data/data/scelist_noATAC.RDS")
scelistnorna <- lapply(scelist, function(sce){
    sce <- swapAltExp(sce, name="ATAC")
    altExp(sce) <- NULL
    sce
})
saveRDS(scelistnorna, file="~/Downloads/multiome_data/data/scelist_noRNA.RDS")
```

# Session Info

```{r}
sessionInfo()
```
