---
title: "Quality Control"
author: "Dario Righelli"
date: "`r Sys.Date()`"
editor_options: 
    chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

# Load libraries and helper functions

This section loads the required R packages for single-cell analysis and sources custom filtering utilities.

```{r libs}
library(SingleCellExperiment)
library(scran)
library(ggpubr)
library(scater)
library(TENxMultiomeTools)
source("~/MultiomeSD/R/filterCells.R")
```

## Load the dataset

Here we load a list of `SingleCellExperiment` objects (samples), either filtered/annotated or unfiltered.

```{r data}
scelist <- readRDS("~/Downloads/multiome_data/data/scelist_filt_anno_dblts_tsne.rds")
#Dropping ATAC for memory consumption
scelist <- lapply(scelist, function(sce){ altExp(sce) <- NULL; sce})
invisible(gc())
# scelist <- readRDS("~/Downloads/multiome_data/data/scelist_unfiltered.RDS")
```

## Define cell types and compute QC metrics

This section collects all subclusters across samples, builds helper vectors of names, and computes QC filtering metrics (RNA + ATAC) for each dataset.

```{r cts}
ct <- unique(unlist(lapply(scelist, function(sce) {unique(sce$subcluster)})))
# not computing metrics during compilation for memory problems 
# scelist <- lapply(scelist, computeSignacMetrics) # > needs ATAC
scelist <- lapply(scelist, computeFilterCellsMetrics, metric="both", highQuant="95%")
```


# Number of cells by metrics

This section summarizes the number of cells passing each quality-control metric (ATAC counts, RNA counts, features, nucleosome signal, TSS enrichment, Signac).

```{r cell_metr, eval=TRUE}
sf <- matrix(nrow=length(scelist), , ncol=7)
for(i in 1:12)
{
    sce <- scelist[[i]]
    tot <- dim(sce)[2]
    natac <- sum(sce$nCount_ATAC_passed)
    nrna <- sum(sce$nCount_RNA_passed)
    nfea <- sum(sce$nFeature_RNA_passed)
    nnucl <- sum(sce$nucleosome_passed)
    ntss <- sum(sce$TSS_passed)
    nsign <- sum(sce$in_signac)
    sf[i,] <- c(tot, natac, nrna, nfea, nnucl, ntss, nsign)
}
rownames(sf) <- names(scelist)
colnames(sf) <- c("Starting", "ATAC_s", "RNA_s", "Feature_s",  "Nucleosome_s", "TSS_s", "Signac")
sfd <- as.data.frame(sf)
knitr::kable(sfd)
```

# Number of cells by metrics and cell types

Here we break down the same QC metrics, but stratified by **cell type** within each sample, to see how filtering affects individual subclusters.

```{r cell_metr_ct, results="asis"}
for (i in 1:12)
{
    sc <- scelist[[i]]
    idxct <- which(unique(sc$subcluster) %in% ct)
    sf <- matrix(nrow=length(idxct), ncol=7)
    k <- 1
    for (j in seq_along(ct))
    {
        idx <- which(sc$subcluster==ct[j])
        if (length(idx)==0)
        {
            print(paste0(names(scelist)[i], " has not any cells for ", ct[j], " cell type\n"))
            next
        }
        sce <- sc[,idx]
        tot <- dim(sce)[2]
        natac <- sum(sce$nCount_ATAC_passed)
        nrna <- sum(sce$nCount_RNA_passed)
        nfea <- sum(sce$nFeature_RNA_passed)
        nnucl <- sum(sce$nucleosome_passed)
        ntss <- sum(sce$TSS_passed)
        nsign <- sum(sce$in_signac)
        sf[k,] <- c(tot, natac, nrna, nfea, nnucl, ntss, nsign)
        k <- k+1
    }
    rownames(sf) <- ct[idxct]
    colnames(sf) <- c("Starting", "ATAC_s", "RNA_s", "Feature_s", "Nucleosome_s", "TSS_s", "Signac")
    sfd <- as.data.frame(sf)
    print(paste0("Sample ", names(scelist)[i], " Cell Type ", ct[j],"\n"))
    print(knitr::kable(sfd))
    cat("\n")
}
```

## Convert logical metadata columns

This section converts all logical (TRUE/FALSE) metadata columns into string labels (“1. TRUE” / “2. FALSE”) for easier tabulation and plotting.

```{r logical}
scelist <- lapply(scelist, function(sce) {
    for(i in 1:dim(colData(sce))[2])
    {
        if( is.logical(colData(sce)[,i]) )
        {
            colData(sce)[,i] <- ifelse(colData(sce)[,i]==TRUE, "1. TRUE", "2. FALSE")
        }
    }
    return(sce)
})
```

# Filtered vs unfiltered cells

This section generates diagnostic plots showing the distribution of cells **before and after filtering**, using either all QC filters or only the Signac filter.

```{r plots, eval=TRUE}
glist <- lapply(seq_along(scelist), function(i)
{
    sce <- scelist[[i]]
    plotFilteredCells(sce, filterCellsBy="custom", customColName="in_signac", name=paste0(names(scelist)[i], " sign_filtered"))
})

lapply(seq_along(scelist), function(i)
{
    sce <- scelist[[i]]
    print(plotFilteredCells(sce, filterCellsBy="all", name=paste0(names(scelist)[i], " unfiltered")))
    print(plotFilteredCells(sce, filterCellsBy="custom", customColName="in_signac", name=paste0(names(scelist)[i], " signac_filtered")))
})
```

# Session Info

```{r}
sessionInfo()
```

