---
title: "Cell Type Annotation"
author: "Elena Zuin"
date: "`r Sys.Date()`"
editor_options: 
    chunk_output_type: console
format:
  html:
    code-fold: true
---

# Introduction

```{r, eval=FALSE}
library(SingleCellExperiment)
library(AllenInstituteBrainData)
library(EnsDb.Mmusculus.v79)
library(scuttle)
library(SingleR)
library(scatter)
```

To identify cell types, we utilized the SingleR annotation method, an automatic tool based on reference data. 

# Reference preparation

The Allen Whole Cortex & Hippocampus - 10x Genomics (v2021) dataset was used as the reference. T
his dataset was imported using the *AllenInstituteBrainData* function from the `AllenInstituteBrainData` package.

```{r, eval=FALSE}
allen_ref <- AllenInstituteBrainData("Allen_Mouse_2021")
rownames(allen_ref) <- rowData(allen_ref)$X
```

The cortical and unwanted subclasses were removed to focus specifically on the hippocampus brain region.

```{r, eval=FALSE}
allen_ref <- allen_ref[, !is.na(allen_ref$subclass_label) &
                           allen_ref$subclass_label != ""]
exclude <- c("V3d", "Meis2", "CR", "Car3", 
             unique(allen_ref$subclass_label[grepl("^L[2-6]", allen_ref$subclass_label)]),
             unique(allen_ref$subclass_label[grepl("^NP", allen_ref$subclass_label)]), 
             unique(allen_ref$subclass_label[grepl("^CT", allen_ref$subclass_label)]))
allen_ref <- allen_ref[, !allen_ref$subclass_label %in% exclude]
```

Gene symbols were mapped to Ensembl Gene IDs using `EnsDb.Mmusculus.v79`, ensuring consistent identifiers for downstream analyses.

```{r, eval=FALSE}
map <- AnnotationDbi::mapIds(EnsDb.Mmusculus.v79, keys = rownames(allen_ref),
                             keytype  = "SYMBOL", column = "GENEID",
                             multiVals = "first")
rowData(allen_ref)$symbol <- rownames(allen_ref)
rowData(allen_ref)$ens <- unname(map)
allen_ref <- allen_ref[!is.na(rowData(allen_ref)$ens), ]
rownames(allen_ref) <- rowData(allen_ref)$ens
```

Since the reference contains around 1 million cells, the dataset was aggregated across all cells belonging to the same subclass, producing one profile per subclass. Counts are log-normalized afterwards.

```{r, eval=FALSE}
allen_subclass <- aggregateAcrossCells(allen_ref, use.assay.type = "counts",
                                       id = DataFrame(label = allen_ref$subclass_label))
allen_subclass <- logNormCounts(allen.subclass)
```

Focusing only on CA1-ProS subclass cells, we further aggregate by cluster label to examine finer-grained organization within this hippocampal population.

```{r, eval=FALSE}
allen_ca1_pros <- allen_ref[, allen_ref$subclass_label == "CA1-ProS"]
allen_ca1_pros_clust <- aggregateAcrossCells(allen_ca1_pros,use.assay.type = "counts",
                                             id = DataFrame(label = allen_ca1_pros$cluster_label)
)
allen_ca1_pros_clust <- logNormCounts(allen_ca1_pros_clust)
```

The processed objects are saved for later reuse.

```{r, eval=FALSE}
save(allen_subclass, allen_ca1_pros_clust,
     file = "allen_10xgenomics_hippo_processed.RData")
```

# Cell-type annotation with SingleR

## Load query dataset

Each query dataset was normalized, and Principal Component Analysis (PCA) and t-SNE were computed for each one.

```{r, eval=FALSE}
multiome <- readRDS("scelist_allexps.RDS")
multiome <- lapply(multiome, function(x) logNormCounts(x))
set.seed(422)
multiome <- lapply(multiome, function(x) runPCA(x, ntop = length(rownames(x))))
set.seed(422)
multiome <- lapply(multiome, function(x) runTSNE(x, dimred = "PCA"))
```

## SingleR

Cell annotation was computed using the *SingleR* function from the `SingleR` package.

```{r, eval=FALSE}
pred_subclass <- lapply(multiome, function(x) SingleR(x, ref = allen_subclass, labels = allen_subclass$subclass_label))
multiome <- lapply(1:length(multiome), function(i) {
  multiome[[i]]$subclass <- pred_subclass[[i]]$labels 
  multiome[[i]]
}
```

For CA1-ProS cells, we annotated the sub-population using the CA1-ProS reference, created previously.

```{r, eval = FALSE}
multiome_ca1_pros <- lapply(multiome, function(x) x[, x$subclass == "CA1-ProS"])
set.seed(422)
multiome_ca1_pros <- lapply(multiome_ca1_pros, function(x) runPCA(x, ntop = length(rownames(x))))
set.seed(422)
multiome_ca1_pros <- lapply(multiome_ca1_pros, function(x) runTSNE(x, dimred = "PCA"))

pred_ca1_pros_subclust <- lapply(multiome_ca1_pros , function(x) SingleR(x, ref = allen_ca1_pros_clust, 
                                                                         labels = allen_ca1_pros_clust$cluster_label))

multiome_ca1_pros <- lapply(1:length(multiome_ca1_pros), function(i) {
  multiome_ca1_pros[[i]]$cluster <- pred_ca1_pros_subclust[[i]]$labels 
  multiome_ca1_pros[[i]]$cluster[which(grepl("-ve", multiome_ca1_pros[[i]]$cluster))] <- "CA1-ve"
  multiome_ca1_pros[[i]]$cluster[which(grepl("-do", multiome_ca1_pros[[i]]$cluster))] <- "CA1-do"
  multiome_ca1_pros[[i]]$cluster[which(grepl("-ProS", multiome_ca1_pros[[i]]$cluster))] <- "CA1-ProS"
  multiome_ca1_pros[[i]]$cluster[which(grepl("_CA1", multiome_ca1_pros[[i]]$cluster))] <- "CA1"
  multiome_ca1_pros[[i]]
}
```

# Visualizing Cell Type Annotations

```{r}
scelist <- readRDS("~/Downloads/multiome_data/data/scelist_SR_dbl_signacfilt_dimred_noATAC_doublrm.RDS")
table(scelist$REP1_S3$labels)

# prendi i conteggi delle labels per ogni campione
label_tables <- lapply(scelist, function(sce) {
  tbl <- table(sce$labels)
  as.data.frame(tbl)
})

# aggiungi il nome del campione
for (nm in names(label_tables)) {
  label_tables[[nm]]$sample <- nm
}

# unisci tutto in un unico data.frame
df <- do.call(rbind, label_tables)

# porta i campioni come colonne, le labels come righe
library(tidyr)
library(dplyr)

label_matrix <- df %>%
  tidyr::pivot_wider(names_from = sample, values_from = Freq, values_fill = 0) %>%
  arrange(Var1)

# rinomina la colonna delle labels
colnames(label_matrix)[1] <- "Label"

knitr::kable(label_matrix)

```

# SessionInfo

```{r}
sessionInfo()
```

