---
title: "Doublets Identification"
author: "Dario Righelli"
date: "`r Sys.Date()`"
editor_options: 
    chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache=TRUE)
```

# Load required libraries

This chunk loads the necessary Bioconductor packages:
- `SingleCellExperiment`: data structure for single-cell assays
- `TENxMultiomeTools`: helpers for 10x Genomics multiome data
- `scater`, `scuttle`: normalization, QC, and visualization tools

```{r}
library(SingleCellExperiment)
library(TENxMultiomeTools)
library(scater)
library(scuttle)
library(ggplot2)
```

# Data loading

## Load preprocessed RNA-only dataset

This chunk loads a preprocessed list of `SingleCellExperiment` objects
from disk. The ATAC modality has been removed, and Signac filtering and
PCA/t-SNE have already been applied.

```{r, cache=FALSE}
scelist <- readRDS("~/Downloads/multiome_data/data/scelist_SR_dbl_signacfilt_dimred_noATAC.RDS")
```

# Preprocess loaded data

## Metadata cleanup and annotation

This chunk (currently not executed) shows how:
- Sample names are assigned to each object
- Redundant metadata is removed
- Conditions and genotypes are annotated based on `sample.id`
- Prediction scores are merged from an external table (`pred`)

```{r, eval=FALSE}
names(scelist) <- lapply(scelist, function(sce){unique(sce$sample.id)})
scelist <- lapply(scelist, function(sce)
{
    colData(sce) <- colData(sce)[, -which(colnames(colData(sce)) %in% 
                                    c("SingleR", "Azimuth", "Azimuth_filteredCells", "SingleR_filteredCells"))]
    colData(sce) <- colData(sce)[, -13] ## 13 is a duplicated barcode col
    sce$condition <- ifelse( length(grep("SD", sce$sample.id))!=0, "SD", "CTRL")
    sce$genotype <- ifelse( length(grep("S3", sce$sample.id))!=0, "S3", "WT")
    
    sce$replicate <- strsplit(x=sce$sample.id[1], "_")[[1]][1]
    sce
})

scelist <- lapply(scelist, function(sce) {sce[,sce$in_signac]})
scelist <- lapply(scelist, function(sce)
{
    cidx <- grep(paste0(unique(sce$sample.id),"$"), rownames(pred))
    predsce <- pred[cidx,]
    rownames(predsce) <- gsub(paste0("_", unique(sce$sample.id)), "", rownames(predsce))
    idx <- match(colnames(sce), rownames(predsce))
    colData(sce) <- cbind(colData(sce), predsce[idx,])
    colnames(colData(sce)) <- gsub("scores", "SRscores", colnames(colData(sce)))
    sce
})
```

## Compute doublets

This chunk performs log-normalization and detects doublets using the
Griffiths method for each dataset (not currently evaluated).

# computing doublets

```{r, eval=FALSE}
library(scuttle)
scelist <- lapply(scelist, logNormCounts)

scelist <- lapply(scelist, function(sce)
{
    computeDoublets(sce, method="griffiths")
})
```

## Dimensionality reduction (PCA and t-SNE)

This chunk runs PCA and then t-SNE on the expression data, projecting
cells into a low-dimensional space for visualization (disabled).

# T-SNE 

```{r, eval=FALSE}
scelist <- lapply(scelist, runPCA)
scelist <- lapply(scelist, runTSNE, dimred="PCA")
```

## Visualize t-SNE colored by cell type labels

This chunk generates t-SNE plots for each dataset, coloring cells by
their assigned cell type label.

```{r tsne}
tsnes <- lapply(scelist, function(sce)
{
    gg <- plotTSNE(sce, color_by="labels") + 
        ggtitle(paste0(sce$replicate, " ", sce$genotype, " ", sce$condition))
    gg 
})
# tsnes
```

# CA1-do

## Extract and visualize CA1-do cells

This chunk filters each dataset to keep only cells annotated as "CA1-do", then:
- Plots their t-SNE embeddings colored by label
- Adds the best score from SingleR predictions (`SRBestScore`)
- Visualizes doublet calls and prediction confidence using a violin plot

```{r ca1do}
scelistdo <- lapply(scelist, function(sce) {sce[,sce$labels=="CA1-do"]})
tsnesdo <- lapply(scelistdo, function(sce)
{
    gg <- plotTSNE(sce, color_by="labels") + 
        ggtitle(paste0(sce$replicate, " ", sce$genotype, " ", sce$condition))
    gg 
})

scelistdo <- lapply(scelistdo, function(sce)
{
    sce$SRBestScore <- sce$SRscores[,1]
    sce
})

vlns <- lapply(scelistdo, function(sce)
{
    gg <- plotColData(sce, x="dbl.calls", y="delta.next", color_by="dbl.scores") + 
        geom_hline(yintercept=0.03)+
        ggtitle(paste0(sce$replicate, " ", sce$genotype, " ", sce$condition))
    gg 
})

idxs3 <- grep("_S3$", names(vlns))
idxs3sd <- grep("_S3SD$", names(vlns))
idxsd <- grep("_SD$", names(vlns))
idxwt <- grep("_WT$", names(vlns))
idxordlist <- c(idxs3, idxs3sd, idxsd, idxwt)
ggpubr::ggarrange(plotlist=vlns[idxordlist], ncol=3)
```

# Filtering doublets + Low SingleR scores

## CA1-do only

## Filter CA1-do cells to remove doublets and low-confidence annotations

This chunk keeps only singlets (non-doublets) and cells with SingleR 
`delta.next > 0.03`.
It then visualizes filtered CA1-do cells using t-SNE.

```{r ca1dof}
scelistdof <- lapply(scelistdo, function(sce)
{
    sce <- sce[,sce$dbl.calls=="singlet"]
    sce[,sce$delta.next>0.03]
})

tsnesdof <- lapply(scelistdof, function(sce)
{
    gg <- plotTSNE(sce, color_by="labels") + 
        ggtitle(paste0(sce$replicate, " ", sce$genotype, " ", sce$condition, " filtered"))
    gg 
})
```

## All other Cell Types 

As done with CA1-do, this chunk filters all datasets:
- Removes doublets (`dbl.calls == "singlet"`)
- Keeps cells with confident label predictions (`delta.next > 0.03`)

It then visualizes the results using t-SNE plots.

```{r allctf}
scelistf <- lapply(scelist, function(sce)
{
    sce <- sce[,sce$dbl.calls=="singlet"]
    sce[,sce$delta.next>0.03]
})

tsnesf <- lapply(scelistf, function(sce)
{
    gg <- plotTSNE(sce, color_by="labels") + 
        ggtitle(paste0(sce$replicate, " ", sce$genotype, " ", sce$condition, " filtered"))
    gg 
})
```

## Combine and compare t-SNE plots

This chunk arranges and displays four t-SNE plots per sample:
1. Original
2. Filtered all cells
3. CA1-do only
4. Filtered CA1-do

```{r plots}
library(ggpubr)
tsnes <- tsnes[idxordlist]
tsnesf <- tsnesf[idxordlist]
tsnesdo <- tsnesdo[idxordlist]
tsnesdof <- tsnesdof[idxordlist]
lapply(seq_along(scelist[idxordlist]), function(i)
{
    ggl <- list(tsnes[[i]], tsnesf[[i]], tsnesdo[[i]], tsnesdof[[i]])
    ggarrange(plotlist=ggl, ncol=2)
})
```

# Summary Table

## Generate summary table of cell type counts

This chunk builds a summary table that compares the number of cells per label
before and after filtering, across all samples. It:
- Constructs a matrix of cell type counts per sample
- Matches labels across filtered and unfiltered data
- Displays the tables using `DT::datatable()` for interactive visualization

```{r dt, cache=FALSE}

nms <- unlist(lapply(scelist, function(sce)
{
    unique(paste0(sce$replicate,  "_", sce$genotype, "_", sce$condition ))
}))

names(nms) <- NULL
lbls <- sort(unique(unlist(lapply(seq_along(scelist[idxordlist]), function(i)
{
    unique(scelist[[i]]$labels)
}))))

scelist <- scelist[idxordlist]
scelistf <- scelistf[idxordlist]
scelistdo <- scelistdo[idxordlist]
scelistdof <- scelistdof[idxordlist]
nms <- nms[idxordlist]

df <- data.frame(labels=lbls)
for(i in seq_along(scelist))
{
    # print(i)
    sn <- nms[i]
    tbls <- table(scelist[[i]]$labels)
    nc <- rep(0, dim(df)[1])
    names(nc) <- df[,1]
    nc[match(names(tbls), names(nc))] <- tbls
    nmdf <- colnames(df)
    df <- cbind.data.frame(df, nc)
    tblsf <- table(scelistf[[i]]$labels)
    nc <- rep(NA, dim(df)[1])
    names(nc) <- df[,1]
    nc[match(names(tblsf), names(nc))] <- tblsf
    df <- cbind.data.frame(df, nc)
    colnames(df) <- c(nmdf, sn, paste0(sn, "_filt"))
}
# df
df <- df[,-1]
dflist <- list()
j=1
for (i in 1:dim(df)[2])
{
    if ((i %% 4) == 0)
    {
        dflist[[j]] <- df[,c((i-3):i)]
        j=j+1
    }
}
DT::datatable(dflist[[1]], class = 'cell-border stripe', options=list("lengthMenu"=c(25,50,100)))
DT::datatable(dflist[[2]], class = 'cell-border stripe', options=list("lengthMenu"=c(25,50,100)))
DT::datatable(dflist[[3]], class = 'cell-border stripe', options=list("lengthMenu"=c(25,50,100)))
DT::datatable(dflist[[4]], class = 'cell-border stripe', options=list("lengthMenu"=c(25,50,100)))
DT::datatable(dflist[[5]], class = 'cell-border stripe', options=list("lengthMenu"=c(25,50,100)))
DT::datatable(dflist[[6]], class = 'cell-border stripe', options=list("lengthMenu"=c(25,50,100)))
```

## Save filtered cell barcodes and align ATAC data accordingly

This chunk (currently disabled) does the following:
- Extracts barcodes from filtered RNA data
- Matches these to corresponding ATAC datasets
- Subsets the ATAC data to include only matching barcodes
- Saves both filtered RNA and ATAC datasets to disk

```{r, eval=FALSE}
bc <- lapply(scelistf, function(sce){
    colnames(sce)
})

saveRDS(scelistf, file="~/Downloads/multiome_data/data/scelist_SR_dbl_signacfilt_dimred_noATAC_doublrm.RDS")


scelistatac <- readRDS("~/Downloads/multiome_data/data/scelist_SR_dbl_signacfilt_dimred_noRNA.RDS")

idx <- match(names(bc), names(scelistatac))
scelistatac<- scelistatac[idx]
scelistatacf<-lapply(seq_along(scelistatac), function(i)
{
    scelistatac[[i]][,bc[[i]]]
})
saveRDS(scelistatacf, file="~/Downloads/multiome_data/data/scelist_SR_dbl_signacfilt_dimred_noRNA_doublrm.RDS")
```


# Session Info

```{r, tidy=TRUE}
sessionInfo()
```




