---
title: "Differential Expression Wild Type for CA1-do"
author: "Dario Righelli"
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

# Loading required libraries

```{r, message=FALSE, warning=FALSE}
library(scater)
library(edgeR)
library(RUVSeq)
library(darioscripts)
library(DT)
library(ggpubr)
source("../R/luciavolcano.R")
library(muscat)
library(biomaRt)

```

# Pseudo-bulk data construction for differential analysis

In this code chunk, we construct **pseudo-bulk RNA-seq data** from single-cell multiome datasets using the `TENxMultiomeTools` package. Specifically, we:

1.  Load a list of `SingleCellExperiment` objects (`scelist`) representing multiple samples.
2.  Aggregate the single-cell count data across cell labels (using `aggregateAcrossCells`) to simulate bulk RNA-seq profiles for each annotated group.
3.  Generate a combined `SingleCellExperiment` object (`scepseudo`) by merging the pseudo-bulk profiles from all samples.
4.  Save the resulting pseudo-bulk dataset to an RDS file for downstream analysis.

This step is performed to prepare the data for **differential expression analysis**, which is more robust when applied to aggregated (pseudo-bulk) counts rather than single-cell counts.

> **Note:** This code chunk is **not evaluated** in the document for time-saving purposes. Instead, we **directly start from the pre-computed results** stored in the `GEX_pseudo_2023.RDS` file.

```{r, eval=FALSE}
scelist <- readRDS("~/Downloads/multiome_data/data/scelist_SR_dbl_signacfilt_dimred_noATAC_doublrm.RDS")

library(TENxMultiomeTools)
scelistpseudo <- lapply(scelist, function(sce)
{
    scea <- aggregateAcrossCells(sce, use.assay.type="counts",
        id=DataFrame(label=sce$labels))
    colnames(scea) <- paste(scea$label, scea$replicate, scea$condition, scea$genotype, sep="_")
    scea
})

scepseudo <- cbind(scelistpseudo[[1]], scelistpseudo[[2]],scelistpseudo[[3]],scelistpseudo[[4]],
                   scelistpseudo[[5]],scelistpseudo[[6]],scelistpseudo[[7]],scelistpseudo[[8]],
                   scelistpseudo[[9]],scelistpseudo[[10]],scelistpseudo[[11]],scelistpseudo[[12]])

saveRDS(scepseudo, file="~/Downloads/multiome_data/data/GEX_pseudo_2023.RDS")

```

# Loading and annotating pseudo-bulk data for differential analysis

In this code chunk, we load the previously computed **pseudo-bulk gene expression dataset** and perform sample- and gene-level annotations in preparation for **differential expression analysis**. The steps include:

1.  **Loading data:**
    -   The pseudo-bulk dataset is read from the RDS file `GEX_pseudo_2023.RDS`.
    -   A composite variable `gcondition` is created by concatenating genotype and condition.
    -   Each sample is annotated with a high-level cell class (`superclass`) based on the `label`, distinguishing non-neuronal, glutamatergic, and GABAergic neuronal populations.
2.  **Annotating negative controls:**
    -   A set of known **negative control genes** is loaded and split into training and testing subsets (90% and 10%, respectively).
    -   Genes in the pseudo-bulk dataset are flagged as negative controls (`negctl`) or test negative controls (`negctltest`) accordingly.
3.  **Annotating positive controls:**
    -   Known **positive control genes** are loaded from an external file and flagged in the dataset (`posctl`).
    -   Their presence in the dataset is also recorded in a `kept` column.
4.  **Filtering samples:**
    -   The dataset is restricted to samples from the **wild-type (WT)** genotype only.

```{r}

multiome <- readRDS("~/Downloads/multiome_data/data/GEX_pseudo_2023.RDS") # pseudobulk


multiome$gcondition <- paste(multiome$genotype, multiome$condition, sep="_")

multiome$superclass <- "Non Neuronal"

multiome$superclass[multiome$label %in% 
        c("CA1-do","CA1-ProS", "CA1", "CA2", "CA3", "SUB-ProS", "DG")] <- "Neuronal - Glutamatergic"
multiome$superclass[multiome$label %in% 
        c("Lamp5", "Sncg", "Vip", "Serpinf1", "Ntng1", "Pvalb", "Sst", "Sst-Chodl")] <- "Neuronal - Gabaergic"


nc <- readRDS("~/Downloads/multiome_data/ctrls/SD_neg_ctrl.RDS")
dim(nc)
rowData(multiome)$negctl <- FALSE
rowData(multiome)$negctl[rownames(multiome) %in% nc$symbol] <- TRUE
rm(.Random.seed, envir=globalenv())
set.seed(3)
idxsamp <- sample(seq_along(nc$ensembl_id), round((dim(nc)[1]*10)/100))
nctest <- nc[idxsamp,]
nc <- nc[-idxsamp,]
rowData(multiome)$negctl <- FALSE
rowData(multiome)$negctl[rownames(multiome) %in% nc$symbol] <- TRUE
rowData(multiome)$negctltest <- FALSE
rowData(multiome)$negctltest[rownames(multiome) %in% nctest$symbol] <- TRUE


newps <- read.table("~/Downloads/multiome_data/ctrls/111022_Positive_Controls_from_BMC_Genomics_AddFile2.txt", header=TRUE)
newpshits <- newps
rowData(multiome)$posctl <- FALSE
rowData(multiome)$posctl[rownames(multiome) %in% newps$Gene_Name] <- TRUE

newpshits$kept <- FALSE
newpshits$kept[which(newpshits$Gene_Name %in% rownames(multiome))] <- TRUE
multiome <- multiome[,multiome$genotype=="WT"]
multiome$condition[multiome$condition=="CTRL"] <- "HC"

```

# Annotating genes with reference metadata

In this code chunk, we enrich the pseudo-bulk dataset with **gene-level annotations** using an external reference table. The steps performed are:

1.  A reference table (`features_duplicates.tsv`) containing gene metadata (such as gene symbols, Ensembl IDs, chromosome locations, and genomic coordinates) is loaded.

2.  Genes in the pseudo-bulk dataset are matched with entries in the reference table based on gene symbols.

3.  For matched genes, the following annotations are added to the `rowData` of the `multiome` object:

    -   `TENxSymbol`: official 10x Genomics gene symbol
    -   `TENxEnsembl`: Ensembl gene ID
    -   `TENxChr`: chromosome
    -   `TENxStart`: transcription start coordinate
    -   `TENxEnd`: transcription end coordinate

This metadata is useful for downstream analysis and interpretation of differential expression results.

```{r}
feat <- read.table("~/Downloads/multiome_data/reference/features_duplicates.tsv", sep="\t", header=TRUE)
rn <- rownames(rowData(multiome)) 
idx <- which(rn %in% feat$symbol)
idx1 <- which(feat$symbol %in% rn)
rowData(multiome)$TENxSymbol[idx] <- feat$symbol[idx1]
rowData(multiome)$TENxEnsembl[idx] <- feat$ensembl[idx1]
rowData(multiome)$TENxChr[idx] <- feat$chr[idx1]
rowData(multiome)$TENxStart[idx] <- feat$start[idx1]
rowData(multiome)$TENxEnd[idx] <- feat$end[idx1]

```

# Exploratory analysis of raw (non-normalized) data

In this code chunk, we explore the structure of the **raw (non-normalized) pseudo-bulk counts** for the `"CA1-do"` cell population using dimensionality reduction techniques:

1.  **Principal Component Analysis (PCA):**
    -   A PCA is computed on the log-transformed raw counts (`log1p(counts(...))`).
    -   The PCA plot distinguishes samples by **condition** using different colors and shapes.
    -   The function `plotPCAs` is used to generate the plot with `cowplot` styling, and color/shape mappings are customized.
2.  **Normalization:**
    -   Log-normalization is then applied using `logNormCounts` normalization to prepare the data for subsequent analyses.

This step provides a visual assessment of variability in the raw data and sets up the object for normalized downstream analyses.

```{r}
ca1do <- multiome[,multiome$label=="CA1-do"]
plotPCAs(log1p(counts(ca1do)),
         colData=colData(ca1do),
         shapeBy="condition",
         colorBy="condition", cowplot=TRUE, prefix="log(raw-counts) CA1-do",
         returnOnly=TRUE, size=3) +
        scale_color_manual(values=c("red", "black")) + scale_shape_manual(values = 16:17) + 
    theme_bw()

colData(ca1do) <- colData(ca1do)[,-which(colnames(colData(ca1do))=="sizeFactor")]
ca1do <- logNormCounts(ca1do)
```

# Data subsetting, upper-quartile normalization, and PCA visualization

First, we apply upper-quartile normalization to account for sequencing depth differences across samples. After normalization, we perform a PCA on the log-transformed normalized counts to visualize how samples cluster by their condition, which helps assess batch effects or biological differences.

```{r}
ca1do <- multiome[, multiome$labels == "CA1-do"]
uq <- EDASeq::betweenLaneNormalization(counts(ca1do), which = "upper")
plotPCAs(log1p(uq),
         colData = colData(ca1do),
         shapeBy = "condition",
         colorBy = "condition",
         cowplot = TRUE,
         prefix = "WT log(UQ)",
         returnOnly = TRUE,
         size = 3) +
  scale_color_manual(values = c("red", "black")) +
  scale_shape_manual(values = 16:17) +
  theme_bw()
```

# RUVs normalization and PCA visualization

To remove unwanted variation not captured by the condition (e.g., batch effects), we apply RUVs normalization using a set of negative control genes. We define sample groups by condition and use k=3 factors to adjust. The normalized counts from RUVs are stored, and a PCA is again performed on these adjusted counts to check if unwanted variation was reduced.

```{r}
grp <- makeGroups(ca1do$condition)
rownames(grp) <- unique(ca1do$condition)

k <- 3
ruvs <- RUVs(x = round(as.matrix(uq)),
             cIdx = rownames(ca1do)[rowData(ca1do)$negctl],
             k = k,
             scIdx = grp,
             round = TRUE)

assays(ca1do)$uqruvsk3 <- ruvs$normalizedCounts
assays(ca1do)$loguqruvsk3 <- log1p(ruvs$normalizedCounts)

plotPCAs(log1p(ruvs$normalizedCounts),
         colData = colData(ca1do),
         shapeBy = "condition",
         colorBy = "condition",
         cowplot = TRUE,
         prefix = paste0("WT log(UQ+RUVsk", k, ")"),
         returnOnly = TRUE,
         size = 3) +
  scale_color_manual(values = c("red", "black")) +
  scale_shape_manual(values = 16:17) +
  theme_bw()
```

# Add RUVs factors to sample metadata and filter genes

We append the unwanted variation factors estimated by RUVs to the sample metadata (`colData`). Next, we prepare for differential expression by creating an `EdgeR` DGEList object and calculating normalization factors. We then filter out genes with low expression, keeping only those with sufficient counts across conditions to ensure robust differential testing.

```{r}
# Add estimated unwanted variation factors to sample metadata
colData(ca1do) <- cbind.DataFrame(colData(ca1do), ruvs$W, returnOnly = TRUE)

# Create DGEList for EdgeR analysis and calculate normalization factors
y <- DGEList(counts(ca1do), samples = colData(ca1do))
y1 <- calcNormFactors(y, method = "upperquartile")

# Filter out lowly expressed genes to improve statistical power
keep <- filterByExpr(y, group = ca1do$condition, min.count = 5, min.total.count = 0)
keep.genes <- data.frame(keep[keep == "TRUE"])
ca1do <- ca1do[rownames(keep.genes), ]
```

# Differential expression analysis with EdgeR and result annotation

We perform differential expression analysis using `EdgeR`, specifying contrasts of interest and incorporating the unwanted variation factors as covariates. After obtaining results, we annotate the differential expression results with gene symbols and Ensembl IDs for interpretability.

```{r}
# Run differential expression analysis with EdgeR using condition contrasts and RUVs factors
cntrs <- c("SD - HC")
reslist <- applyEdgeR(counts = counts(ca1do),
                      colData = colData(ca1do),
                      factors = "condition",
                      wnames = colnames(ruvs$W),
                      contrasts = cntrs,
                      p.threshold = 1,
                      edgeroffs = log(y1$samples$lib.size * y1$samples$norm.factors))

# Annotate DE results with gene symbols and Ensembl IDs
newpshitsl <- newpshits[, c(2, 1)]
reslist <- lapply(reslist, function(r) {
  r$gene <- rownames(r)
  r$ensembl <- rowData(ca1do)$TENxEnsembl[match(r$gene, rownames(rowData(ca1do)))]
  return(r)
})
```

# Plotting results and summarizing DEGs

We visualize the differential expression results with volcano plots highlighting positive controls, and histograms of p-values to check statistical distributions. We then summarize the number of significant DEGs and count how many positive and negative control genes are detected as differentially expressed.

Then, we add logical flags to indicate which genes are significantly differentially expressed and which correspond to positive control genes. Finally, we attach the full differential expression results back to the dataset’s row metadata for further use or export.

```{r}
# Volcano plot to visualize DEGs and positive control genes
luciaVolcanoPlot1(reslist[[1]],
                  positive.controls.df = newpshitsl,
                  prefix = paste0(names(reslist)[1]),
                  threshold = 0.05) + theme_bw()

# Histogram of p-values to check distribution of significance
PlotHistPvalPlot(reslist[[1]],
                 prefix.plot = names(reslist)[1],
                 show.plot.flag = FALSE) + theme_bw() +
  ggtitle("SD-HC P-Value Histogram")

# Summarize number of significant DEGs
DEGsk3 <- unlist(lapply(reslist, function(r) sum(r$FDR < 0.05)))

# Count positive controls detected among significant DEGs
pcDEGsk3 <- unlist(lapply(reslist, function(r) {
  r <- r[r$FDR < 0.05, ]
  sum(tolower(rownames(r)) %in% tolower(newpshits$Gene_Name))
}))

# Count negative controls detected among significant DEGs
ncr3d <- unlist(lapply(reslist, function(r) {
  r <- r[r$FDR < 0.05, ]
  sum(tolower(rownames(r)) %in% tolower(nc$symbol))
}))

# Flag significant DEGs and mark positive control genes
reslist <- lapply(reslist, function(r) {
  r$sign <- FALSE
  r$sign[r$FDR < 0.05] <- TRUE
  r$poscont <- FALSE
  r$poscont[tolower(rownames(r)) %in% tolower(newpshits$Gene_Name)] <- TRUE
  r
})

# Attach DE results to the dataset's rowData metadata
r3rsuq <- reslist[[1]]
ca1do <- ca1do[match(rownames(r3rsuq), rownames(ca1do)), ]
rowData(ca1do)$edger <- r3rsuq

```

# Session Info

```{r, tidy=TRUE}
sessionInfo()
```
