---
title: "Multi-Omic Factor Analysis (MOFA+) RNA"
author: "Dario Righelli"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```


# utils

```{r}
getTopWeightsMOFA2 <- function(model, factor, nfeatures=20, scale=TRUE, abs=FALSE)
{
    library(dplyr)
    abs=TRUE
    W <- get_weights(model, factors = factor, as.data.frame=TRUE)
    if (scale) W$value <- W$value/max(abs(W$value))
    W <- W[W$value != 0, ]
    W$sign <- ifelse(W$value > 0, "+", "-")
    if (abs) W$value <- abs(W$value)
    W <- W[with(W, order(-abs(value))), ]
    W <- as.data.frame(top_n(group_by(W, factor), n = nfeatures,
                             wt = value))

    W$feature_id <- W$feature
    W$feature_id <- factor(W$feature_id, levels = rev(unique(W$feature_id)))
    return(W)
}
```

# Loading Libraries

```{r, message=FALSE}
library(MOFA2)
library(MultiAssayExperiment)
library(scater)
library(scran)
library(Matrix)
library(dplyr)
```

# preprocessing data

## preprocessing RNA

```{r}
set.seed(999)
scelistrna <- readRDS("~/Downloads/multiome_data/data/scelist_SR_dbl_signacfilt_dimred_noATAC_doublrm.RDS")
scelistrnawt <- scelistrna[-grep("S3", names(scelistrna))]
mornawt <- do.call(cbind,scelistrnawt)
mornawtca1do <- mornawt[,mornawt$labels=="CA1-do"]
hvg <- getTopHVGs(mornawtca1do, n=1000)
mornawtca1dohvg <- mornawtca1do[hvg,]
mornawtca1dohvg$sample.id <- gsub("WT", "HC", mornawtca1dohvg$sample.id)
colnames(mornawtca1dohvg) <- paste0(mornawtca1dohvg$sample.id, "_", colnames(mornawtca1dohvg))
mornawtca1dohvg$condition <- gsub("CTRL", "HC", mornawtca1dohvg$condition)
```


# Preparing mofa object

we create a MultiAssayExperiment and then a MOFA object without groups

```{r, warning=FALSE}
explistwt <- list("RNA"=mornawtca1dohvg)
(mae <- MultiAssayExperiment(experiments=explistwt))


sampleMap(mae)$condition <- unlist(lapply(strsplit(as.character(sampleMap(mae)$colname), "_"), function(x) {x[2]}))
sampleMap(mae)$rep <- unlist(lapply(strsplit(as.character(sampleMap(mae)$colname), "_"), function(x) {x[1]}))
rep <- unlist(lapply(strsplit(unique(unlist(colnames(mae))),"_"), function(x) {x[1]}))
cond <- unlist(lapply(strsplit(unique(unlist(colnames(mae))),"_"), function(x) {x[2]}))
cd <- data.frame(cell=unique(unlist(colnames(mae))), replicate=rep, condition=cond, rep_cond=paste(rep, cond, sep="_"))
rownames(cd) <- cd$cell
colData(mae) <- DataFrame(cd)

mofa <- create_mofa_from_MultiAssayExperiment(mae)#, groups = "replicate")

data_opts <- get_default_data_options(mofa)
data_opts$scale_views <- TRUE
data_opts$scale_groups <- TRUE

model_opts <- get_default_model_options(mofa)
model_opts$num_factors <- 10

train_opts <- get_default_training_options(mofa)

MOFAobject <- prepare_mofa(
  object = mofa,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
plot_data_overview(MOFAobject)
```

# Training MOFA

We train a model based on the previous object.

```{r, message=FALSE, results='hide'}
outfile = file.path("./model_RNA.hdf5")
MOFAobject.trained <- run_mofa(MOFAobject, outfile, use_basilisk=TRUE)
model <- MOFAobject.trained
sm <- samples_metadata(model)
cd$sample <- cd$cell
sm$cell <- sm$sample
newcm <- merge(sm, cd, by.x="sample", by.y="sample")
samples_metadata(model) <- newcm
```

Plotting the factors to understand if there are some catching the data
variability between the conditions

```{r, message=FALSE}
plot_factors(model,
             factors = 1:10,
             color_by = "condition"
)
```

We notice that factor 3 and 4 catch the differences between the conditions

```{r}
plot_factors(model,
             factors = c(3:4),
             color_by = "condition"
)
```

Inspecting the genes contributing to the explained variability by the factors

```{r}
plot_weights(model,
         factor = c(3),
         nfeatures = 20,     # Number of features to highlight
         scale = TRUE,          # Scale weights from -1 to 1
         abs = FALSE             # Take the absolute value?
)

plot_weights(model,
         factor = c(4),
         nfeatures = 20,     # Number of features to highlight
         scale = TRUE,          # Scale weights from -1 to 1
         abs = FALSE             # Take the absolute value?
)

```

Plotting top 10 genes for each factor

```{r}
plot_top_weights(model,
           factors = c(3),
           nfeatures = 20,     # Number of features to highlight
           scale = TRUE,          # Scale weights from -1 to 1
           abs = FALSE             # Take the absolute value?
)

plot_top_weights(model,
           factors = c(4),
           nfeatures = 20,     # Number of features to highlight
           scale = TRUE,          # Scale weights from -1 to 1
           abs = FALSE             # Take the absolute value?
)


# write.table(x = W, file="top_100_MOFA_RNA_Factors34.tsv", row.names = FALSE, quote=FALSE, sep = "\t")

for(f in c(1:10))
{
    W <- getTopWeightsMOFA2(model, factor = f, nfeatures = 100)
    fc <- paste0("Factor", f)
    write.table(x = W[W$factor==fc,], file=paste0("top_100_MOFA_RNA_",fc, ".tsv"), row.names = FALSE, quote=FALSE, sep = "\t")
}

```

Computing and inspecting the tSNE based on the first 5 MOFA factors

```{r}

set.seed(999)
model <- run_tsne(model, factors = 1:5)#, perplexity=100)

tsne <- model@dim_red$TSNE
rownames(tsne) <- tsne$sample
tsne <- as.matrix(tsne[,-1])
cd <- model@samples_metadata
sce <- SingleCellExperiment(
  assays = list(),  # Inserisci qui eventuali matrici di espressione, se disponibili
  colData = cd
)

# Aggiungi la riduzione dimensionale TSNE all'oggetto SCE
reducedDims(sce)$TSNE <- tsne

library(scater)

plotTSNE(sce, colour_by="condition") +
  scale_color_manual(name="Condition", values = c("HC" = "#F8766D", "SD" = "#00BFC4"))
```

# Session Info

```{r}
sessionInfo()
```

