---
title: "Multi-Omic Factor Analysis (MOFA+) RNA"
author: "Dario Righelli"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, eval=TRUE)
```

# Additional util

Here we set a function specifically designed to extract the a user-defined number of features from the MOFA+ weights present in the model.

```{r}
getTopWeightsMOFA2 <- function(model, factor, nfeatures=20, scale=TRUE, abs=FALSE)
{
    library(dplyr)
    abs=TRUE
    W <- get_weights(model, factors = factor, as.data.frame=TRUE)
    if (scale) W$value <- W$value/max(abs(W$value))
    W <- W[W$value != 0, ]
    W$sign <- ifelse(W$value > 0, "+", "-")
    if (abs) W$value <- abs(W$value)
    W <- W[with(W, order(-abs(value))), ]
    W <- as.data.frame(top_n(group_by(W, factor), n = nfeatures,
                             wt = value))
    
    W$feature_id <- W$feature
    W$feature_id <- factor(W$feature_id, levels = rev(unique(W$feature_id)))
    return(W)
}
```

# Loading Libraries

```{r, message=FALSE}
library(MOFA2)
library(MultiAssayExperiment)
library(scater)
library(scran)
library(Matrix)
library(dplyr)
```

# Preprocessing and Selection of Highly Variable Genes in CA1-do Cells

We load single-cell RNA-seq data and then we merge the selected wild-type samples into a single object and focus on cells labeled as "CA1-do". We identify the top 1,000 highly variable genes and subset the data accordingly.

```{r, eval=FALSE}
set.seed(999)
scelistrna <- readRDS("~/Downloads/multiome_data/data/scelist_SR_dbl_signacfilt_dimred_noATAC_doublrm.RDS")
scelistrnawt <- scelistrna[-grep("S3", names(scelistrna))]
mornawt <- do.call(cbind,scelistrnawt)
mornawtca1do <- mornawt[,mornawt$labels=="CA1-do"]
hvg <- getTopHVGs(mornawtca1do, n=1000)
mornawtca1dohvg <- mornawtca1do[hvg,]
mornawtca1dohvg$sample.id <- gsub("WT", "HC", mornawtca1dohvg$sample.id)
colnames(mornawtca1dohvg) <- paste0(mornawtca1dohvg$sample.id, "_", colnames(mornawtca1dohvg))
mornawtca1dohvg$condition <- gsub("CTRL", "HC", mornawtca1dohvg$condition)
```

# Preparing the MOFA Model from Single-Modality RNA Data

We construct a `MultiAssayExperiment` object from the RNA expression matrix and extract replicate and condition information from column names. This metadata is added to the sample annotations. We then create a MOFA object without specifying groups, configure the data, model, and training options—scaling both views and groups, and setting the number of factors to 10. Finally, we prepare the model and visualize a summary of the input data structure.

```{r, warning=FALSE, eval=FALSE}
explistwt <- list("RNA"=mornawtca1dohvg)
(mae <- MultiAssayExperiment(experiments=explistwt))
sampleMap(mae)$condition <- unlist(lapply(strsplit(as.character(sampleMap(mae)$colname), "_"), function(x) {x[2]}))
sampleMap(mae)$rep <- unlist(lapply(strsplit(as.character(sampleMap(mae)$colname), "_"), function(x) {x[1]}))
rep <- unlist(lapply(strsplit(unique(unlist(colnames(mae))),"_"), function(x) {x[1]}))
cond <- unlist(lapply(strsplit(unique(unlist(colnames(mae))),"_"), function(x) {x[2]}))
cd <- data.frame(cell=unique(unlist(colnames(mae))), replicate=rep, condition=cond, rep_cond=paste(rep, cond, sep="_"))
rownames(cd) <- cd$cell
colData(mae) <- DataFrame(cd)
mofa <- create_mofa_from_MultiAssayExperiment(mae)#, groups = "replicate")
data_opts <- get_default_data_options(mofa)
data_opts$scale_views <- TRUE
data_opts$scale_groups <- TRUE
model_opts <- get_default_model_options(mofa)
model_opts$num_factors <- 10
train_opts <- get_default_training_options(mofa)

MOFAobject <- prepare_mofa(
  object = mofa,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
plot_data_overview(MOFAobject)
```

# Training the MOFA Model

We train the MOFA model using the previously configured object and save the output to an HDF5 file. After training, we extract the sample metadata and merge it with the original cell annotations to ensure consistent labeling. The updated metadata is then assigned back to the trained model.

```{r, message=FALSE, results='hide', eval=FALSE}
outfile = file.path("./model_RNA.hdf5")
MOFAobject.trained <- run_mofa(MOFAobject, outfile, use_basilisk=TRUE)
model <- MOFAobject.trained
sm <- samples_metadata(model)
cd$sample <- cd$cell
sm$cell <- sm$sample
newcm <- merge(sm, cd, by.x="sample", by.y="sample")
samples_metadata(model) <- newcm
```

# Visualizing Latent Factors by Condition

We plot the first ten latent factors inferred by the MOFA model, coloring the samples by condition. This allows us to assess whether any of the factors capture variability associated with the experimental conditions.

```{r, message=FALSE}
model <- load_model("~/Downloads/multiome_data/data/model_RNA.hdf5")
samples_metadata(model)$condition <- "HC"
samples_metadata(model)$condition[grep("SD", samples_metadata(model)$sample)] <- "SD"
plot_factors(model,
             factors = 1:10,
             color_by = "condition"
)
```

# Highlighting Factors Capturing Condition Differences

We focus on factors 3 and 4, which clearly distinguish between the conditions, confirming their relevance in capturing the underlying biological variability related to the experimental groups.

```{r}
plot_factors(model,
             factors = c(3:4),
             color_by = "condition"
)
```

# Exploring Gene Contributions to Key Factors

We examine the top 20 genes with the highest weighted contributions to factors 3 and 4. The weights are scaled between -1 and 1 to highlight genes that drive the variability captured by these factors, providing insight into the molecular features associated with condition differences.

```{r}
plot_weights(model,
         factor = c(3),
         nfeatures = 20,     # Number of features to highlight
         scale = TRUE,          # Scale weights from -1 to 1
         abs = FALSE             # Take the absolute value?
)

plot_weights(model,
         factor = c(4),
         nfeatures = 20,     # Number of features to highlight
         scale = TRUE,          # Scale weights from -1 to 1
         abs = FALSE             # Take the absolute value?
)

```

# Visualizing Top Gene Weights for Selected Factors

We display the top 20 genes contributing to factors 3 and 4 using scaled weights. This visualization helps identify key genes driving the differences between conditions captured by these latent factors.

```{r}
plot_top_weights(model,
           factors = c(3),
           nfeatures = 20,     # Number of features to highlight
           scale = TRUE,          # Scale weights from -1 to 1
           abs = FALSE             # Take the absolute value?
)

plot_top_weights(model,
           factors = c(4),
           nfeatures = 20,     # Number of features to highlight
           scale = TRUE,          # Scale weights from -1 to 1
           abs = FALSE             # Take the absolute value?
)
```

# Exporting Top Gene Weights for All Factors

Thanks to our util, we can iterate through factors 1 to 10, extracting the top 100 genes contributing to each factor. Then we can save the results as separate tab-delimited files for downstream analysis or reporting.

```{r, eval=FALSE}

for(f in c(1:10))
{
    W <- getTopWeightsMOFA2(model, factor = f, nfeatures = 100)
    fc <- paste0("Factor", f)
    write.table(x = W[W$factor==fc,], file=paste0("top_100_MOFA_RNA_",fc, ".tsv"), row.names = FALSE, quote=FALSE, sep = "\t")
}

```

# Computing and Visualizing t-SNE on MOFA Factors

We compute a t-SNE embedding using the first five MOFA latent factors to capture the main sources of variability in a reduced dimensional space. The resulting coordinates are incorporated into a `SingleCellExperiment` object and plotted, coloring cells by condition to visualize clustering patterns and potential separation between groups.

```{r}

set.seed(999)
model <- run_tsne(model, factors = 1:5)

tsne <- model@dim_red$TSNE
rownames(tsne) <- tsne$sample
tsne <- as.matrix(tsne[,-1])
cd <- model@samples_metadata
sce <- SingleCellExperiment(
  assays = list(),  
  colData = cd
)

reducedDims(sce)$TSNE <- tsne

library(scater)

plotTSNE(sce, colour_by="condition") +
  scale_color_manual(name="Condition", values = c("HC" = "#F8766D", "SD" = "#00BFC4"))
```

# Session Info

```{r}
sessionInfo()
```
